---
title: "Optimising random forest for variable selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Optimising random forest for variable selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## Introduction 

Random forest is a widely used machine learning technique, particularly effective for tasks such as variable selection, where the importance of each variable is assessed and the most significant variables are chosen. Despite its numerous advantages, it is important to recognize that random forest is inherently non-deterministic. This means that repeating the analysis, even with the same data set, can lead to variations in variable importance estimates due to the stochastic nature of the algorithm.

To demonstrate the functionality of the `optRF` package, we will use the `SNPdata` data set included in the package.  This data set contains in the first column the yield of 250 wheat individuals as well as 5000 genomic markers (so called SNP markers) that can contain either the value 0 or 2. While this example focuses on genomic research, the `optRF` package is versatile and can optimize random forest for any type of data set. Before proceeding, ensure the necessary packages are loaded into your R environment. In addition to `optRF`, the `ranger` package (for efficient random forest implementation) and `minpack.lm` package (for optimization tasks) are required:

```{r message=FALSE, warning=FALSE}
library(ranger)
library(minpack.lm)
library(optRF)
```

## Variable importance estimation using random forest
Random forest is a powerful tool for estimating the importance of predictor variables (in this case the genomic markers) based on their association with a response variable (in this case the yield). In the `ranger` function, variable importance can be calculated using the `importance` argument, which can be set to `"impurity"`, `"impurity_corrected"`, or `"permutation"`. In this example, we define variable importance using the `"permutation"` method. To ensure reproducibility, we set a seed before running the analysis:

```{r message=FALSE}
set.seed(123) # Set a seed for reproducibility
RFmodel = ranger(y=SNPdata[,1], x=SNPdata[,-1], write.forest = TRUE, importance="permutation")
D_VI = data.frame(variable = names(SNPdata)[-1], importance = RFmodel$variable.importance)
D_VI = D_VI[order(D_VI$importance, decreasing=TRUE),]
```

The data frame `D_VI` now contains the variable importance scores for each predictor, sorted in descending order. These scores provide insights into the relative significance of each variable in the data set. Since random forest is a non-deterministic method, repeated analyses using the same data set can yield different variable importance scores. To illustrate this, we repeat the analysis with a different seed:

```{r message=FALSE}
set.seed(321) # Set a seed for reproducibility
RFmodel_2 = ranger(y=SNPdata[,1], x=SNPdata[,-1], write.forest = TRUE, importance="permutation")
D_VI_2 = data.frame(variable = names(SNPdata)[-1], importance = RFmodel_2$variable.importance)
D_VI_2 = D_VI_2[order(D_VI_2$importance, decreasing=TRUE),]
```

We can now merge `D_VI` and `D_VI_2` to see how much the variable importance estimates varied simply by repeating random forest after setting a different seed: 

```{r fig.width=6, fig.height=4.5, fig.align='center'}
M = merge(D_VI, D_VI_2, by="variable")
plot(M$importance.x, M$importance.y,
     xlab="Variable importances in the first run", ylab="Variable importances in the second run")
cor(M$importance.x, M$importance.y)
```

The scatter plot and the correlation coefficient (0.27) reveal substantial variability in the variable importance estimates between the two runs. This demonstrates how the non-deterministic nature of random forest can lead to inconsistencies in variable rankings.

## Increasing the stability of random forest

Random forest operates by growing multiple decision trees (or regression trees, in this case) and calculating variable importance within each tree. Since each tree is built independently, the importance values for a given variable can differ between trees. To provide a more reliable measure, random forest averages these importance values across all trees. Consequently, the stability of the variable importance estimates improves as the number of trees increases.

By default, R packages like `ranger`, `randomForest`, and `Rborist` typically use 500 trees. However, for the `SNPdata` data set, 500 trees may not be sufficient to achieve stable variable importance estimates and reliable selections.

To evaluate how the number of trees affects both variable importance stability (correlation of variable importance estimates across runs) and computation time, we tested random forests with different tree counts: 500, 2,500, 5,000, 10,000, 15,000, and 20,000 trees. The stability was calculated as the correlation between variable importance scores obtained from two repeated runs of random forest, each using a different seed. Additionally, we measured the average computation time for generating these random forests.

```{r message=FALSE}
num.trees_values = c(500, 2500, 5000, 10000, 15000, 20000)
result = data.frame()
for(i in num.trees_values){

  start.time = Sys.time()

  set.seed(123)
  RFmodel_1 = ranger(y=SNPdata[,1], x=SNPdata[,-1], num.trees=i,
                     write.forest = TRUE, importance="permutation")
  D_VI_1 = data.frame(variable = names(SNPdata)[-1], importance = RFmodel_1$variable.importance)

  set.seed(321)
  RFmodel_2 = ranger(y=SNPdata[,1], x=SNPdata[,-1],  num.trees=i,
                     write.forest = TRUE, importance="permutation")
  D_VI_2 = data.frame(variable = names(SNPdata)[-1], importance = RFmodel_2$variable.importance)
  end.time = Sys.time()

  M = merge(D_VI_1, D_VI_2, by="variable")
  result = rbind(result, data.frame(number_of_trees = i,
                                    variable_importance_stability = cor(M$importance.x, M$importance.y),
                                    computation_time = (end.time - start.time)/2))
}
result
```

As expected, the stability of variable importance estimates improves with a higher number of trees. For example, stability rises from 0.27 at 500 trees to 0.92 at 20,000 trees.
However, the relationship between the number of trees and variable importance stability is nonlinear. Initially, stability increases rapidly with additional trees but eventually begins to plateau. This diminishing return indicates that beyond a certain number of trees, the stability gains become marginal. In contrast to stability, computation time increases linearly with the number of trees. This is a predictable outcome since each additional tree contributes directly to processing time. We can plot the stability and computation time to better understand these trends:

```{r fig.width=8, fig.height=4.5, fig.align='center'}
par(mfrow=c(1,2))
plot(variable_importance_stability ~ number_of_trees, data=result)
plot(computation_time ~ number_of_trees, data=result)
abline(lm(result$computation_time ~ result$number_of_trees), lwd=2, col="grey")
```

Using the observed data points, it is now possible to model the relationship between the variable importance stability (s) and the number of trees (t) using a two parameter logistic regression (2PL) model 
$$
\hat s_j = \frac{1}{1+\left( \frac{\theta_1}{t_j}\right)^{\theta_2}}
$$ 
where $\theta_1$ denotes the number of trees at which the variable importance stability reaches 0.5 and $\theta_2$ denotes the slope of the curve at $\theta_1$, indicating the steepness of the stability increase. 

To estimate the parameters $\theta_1$ and $\theta_2$, we use the `nlsLM` function from the `minpack.lm` package. This function requires reasonable starting values for the parameters. We can assume that 1000 trees is a realistic value for $\theta_1$ and that the slope might be 1 at this point of the curve. 

```{r message=FALSE}
stability_model = nlsLM(variable_importance_stability ~ 1 / (1+(theta1/number_of_trees)^theta2), data=result,
                        start=c(theta1=1000, theta2=1),
                        control = nls.lm.control(maxiter = 1024))
coefficients(stability_model)
```


Now that the `nlsLM` function has defined certain values for $\theta_1$ and $\theta_2$, the model can be used to estimate the variable importance stability for any number of trees:

```{r fig.width=6, fig.height=4.5, fig.align='center'}
TwoPLmodel = function(number_of_trees, theta1, theta2){
  1 / (1+(theta1/number_of_trees)^theta2)
}
estimated_stability = data.frame(number_of_trees = seq(10, 100000, 10), 
                                 variable_importance_stability = TwoPLmodel(seq(10, 100000, 10), 
                                                                            coefficients(stability_model)[1], 
                                                                            coefficients(stability_model)[2]))
plot(estimated_stability$variable_importance_stability ~ estimated_stability$number_of_trees,
     type="l", col="navyblue", lwd=3, xlab="number_of_trees", ylab="variable_importance_stability")
```

One can see that the variable importance stability increases rapidly with the number of trees at first but levels off as the tree count grows. This diminishing return suggests that adding more trees beyond a certain point has minimal impact on variable importance stability. Since computation time increases linearly with the number of trees, an optimal trade-off point can be identified. For instance, we could define the optimal number of trees as the point where adding additional trees increases variable importance stability by only $10^{-6}$ or less. 


## Using optRF to optimise random forest for variable importance estimation

The optimal number of trees in a random forest model is specific to the data set being analyzed. Therefore, it must be estimated for each new data set to achieve stable variable importance estimates and reliable results. The `optRF` package provides a streamlined approach to determine the optimal number of trees automatically. Therefore, it performs the calculations as shown above with the difference that random forest is not only repeated twice but ten times to ensure more stable estimates of the variable importance stability. It calculates variable importance stability using the intraclass correlation between the results of these repetitions. Besides this, all the code above can be run by using the `opt_importance` function from the `optRF` package. To use it, the response variable from the training data set is passed to the `y` argument and the predictor variables from the training data set are provided in the `X` argument:

```{r message=FALSE, warning=FALSE}
set.seed(123)
optRF_result = opt_importance(y=SNPdata[,1], X=SNPdata[,-1])
```

Once the `opt_importance` function has completed, the recommended number of trees and the expected stability of the random forest model can be accessed as follows:

```{r message=FALSE}
optRF_result$recommendation
optRF_result$expected_RF_stability
```

Here, the optimal number of trees to receive stable variable importance estimates is 40,000 trees. Using random forest with this number of trees, one can expect a variable importance stability of 0.96. 

### The effect of the optimal number of trees
Now that the optimal number of trees has been determined, we can repeat the analysis from above to see how the stability of random forest changed when running random forest with 40,000 trees:


```{r message=FALSE, fig.width=6, fig.height=4.5, fig.align='center'}
set.seed(123)
RFmodel_1 = ranger(y=SNPdata[,1], x=SNPdata[,-1], num.trees=optRF_result$recommendation,
                 write.forest = TRUE, importance="permutation")
D_VI_1 = data.frame(variable = names(SNPdata)[-1], 
                  importance = RFmodel_1$variable.importance)

set.seed(321)
RFmodel_2 = ranger(y=SNPdata[,1], x=SNPdata[,-1], num.trees=optRF_result$recommendation,
                   write.forest = TRUE, importance="permutation")
D_VI_2 = data.frame(variable = names(SNPdata)[-1], 
                    importance = RFmodel_2$variable.importance)

M = merge(D_VI_1, D_VI_2, by="variable")
plot(M$importance.x, M$importance.y,
     xlab="Variable importances in the first run", ylab="Variable importances in the second run")
cor(M$importance.x, M$importance.y)
```

With 40,000 trees, the correlation between the variable importance estimates in the two repetitions is 0.96, a dramatic improvement over the correlation of 0.27 observed with 500 trees. The scatter plot of variable importance estimates from the two runs shows a tightly clustered pattern, confirming consistency in the results and the high correlation aligns with the stability value predicted by `opt_importance`, demonstrating the reliability of the function's recommendation.


## Using optRF to optimise random forest for variable selection
With stable variable importance estimates obtained, these can now be utilized to identify and select the most important variables from the data set. The selection process first involves defining a threshold for the number of variables to be selected. Subsequently, in each repetition of random forest, the variables with the highest variable importance estimates can be classified as `selected` and the rest of the variables as `rejected`. Fleiss’ Kappa is a suitable metric for quantifying the stability of the variable selection process, as it measures agreement among these categorizations in repeated analyses.

The number of variables selected should align with the number of truly important variables in the data set. Instead of selecting an arbitrary number, this should be data-driven. By default, `opt_importance` will determine the optimal number of trees to receive stable selections when selecting the top 5% of variables from the data set. With this data set, this would mean to select for the 250 most important variables in the data set. However, the assumption that 250 important variables exist in the data set may not actually hold. 

To better estimate the number of genuinely important variables in the data set, random forest can be run with the number of trees that was recommended by the `opt_importance` function to get stable variable importance estimates, which can be visualised using a histogram:

```{r message = FALSE, warning=FALSE, fig.width=6, fig.height=4.5, fig.align='center'}
set.seed(123)
RFmodel = ranger(y=SNPdata[,1], x=SNPdata[,-1], num.trees=optRF_result$recommendation,
                   write.forest = TRUE, importance="permutation")
hist(RFmodel$variable.importance, 
     main="Histogram of variable importances", xlab="")
```

One can see that the majority of variables have a variable importance between -5 and 5. Only a small subset of variables has importance scores above 5, likely indicating genuinely important predictors, which should be selected. To refine the selection process, we can calculate the number of variables with variable importance above 5: 

```{r fig.width=6, fig.height=4.5, fig.align='center'}
hist(RFmodel$variable.importance, ylim=c(0,20), 
     main="Histogram of variable importances", xlab="")
abline(v=5, col="red", lwd=4)
selection.size = sum(RFmodel$variable.importance>5)
selection.size
```

In this data set, 39 variables have a variable importance above the threshold of 5. So, we can rerun the `opt_importance` function and this time specify that the 39 most important variables should be selected. This is achieved through the `alpha` argument, which allows for either an absolute number of variables or a relative proportion of variables to be selected from the data set:

```{r message = FALSE, warning=FALSE}
set.seed(123)
optRF_result = opt_importance(y=SNPdata[,1], X=SNPdata[,-1], 
                              recommendation = "selection",
                              alpha = selection.size)
```

The results from this iteration of `opt_importance` provide an updated recommendation for the optimal number of trees required to achieve stable variable selection when focusing on the 39 most important variables. The output reveals the key metrics, including the optimal number of trees, the expected stability of variable importance, the stability of variable selection, and the estimated runtime:

```{r message=FALSE}
optRF_result$recommendation
optRF_result$expected_RF_stability
```

The results indicate that 42,000 trees are required to achieve stable selection of the 39 variables, resulting in a variable importance stability of 0.96 and a selection stability of 0.94. These metrics demonstrate significant improvement in stability compared to prior analyses with fewer trees.

With the optimized number of trees determined, the analysis can be repeated to evaluate how consistently the same variables are selected across two independent random forest runs. The process involves running random forest twice, using the recommended 42,000 trees, and recording the top 39 variables selected in each run based on their importance rankings:

```{r message = FALSE, warning=FALSE}
set.seed(123)
RFmodel_1 = ranger(y=SNPdata[,1], x=SNPdata[,-1], num.trees=optRF_result$recommendation,
                   write.forest = TRUE, importance="permutation")
D_VI_1 = data.frame(variable = names(SNPdata)[-1],
                    importance = RFmodel_1$variable.importance)
D_VI_1 = D_VI_1[order(D_VI_1$importance, decreasing=TRUE),]
D_VI_1$variable_ranking = c(1:nrow(D_VI_1))
selected_variables_1 = D_VI_1[1:selection.size,1]

set.seed(321)
RFmodel_2 = ranger(y=SNPdata[,1], x=SNPdata[,-1], num.trees=optRF_result$recommendation,
                   write.forest = TRUE, importance="permutation")
D_VI_2 = data.frame(variable = names(SNPdata)[-1],
                    importance = RFmodel_2$variable.importance)
D_VI_2 = D_VI_2[order(D_VI_2$importance, decreasing=TRUE),]
D_VI_2$variable_ranking = c(1:nrow(D_VI_2))
selected_variables_2 = D_VI_2[1:selection.size,1]

sum(selected_variables_1 %in% selected_variables_2)
```

In this analysis, 36 out of the 39 selected variables overlap between the two repeated runs of random forest with the recommended number of trees which translates to 92% of the variables to be selected in a stable way.  


## Fine-tuning the optimal number of trees for variable selection

This approach effectively estimates the number of important variables in the data set and determines the optimal number of trees necessary to achieve stable variable selection. However, it also highlights that even with optimization, some variability persists in the selected variables across repeated random forest runs. Nonetheless, with the relationship between stability and the number of trees now established, additional functions from the `optRF` package can be utilized to further refine the analysis, including estimating computation time specific to the machine used for the `opt_importance` function.

The relationship between stability and the number of trees can be conveniently visualized using the `plot_stability` function. This visualization allows users to specify the range of tree counts to consider, the stability measure of interest (variable importance stability or selection stability), and whether the recommended number of trees should be highlighted. For instance, visualizing selection stability from 0 to 100,000 trees without indicating the recommended number of trees can be achieved as follows:

```{r fig.width=6, fig.height=4.5, fig.align='center'}
plot_stability(optRF_result, measure="selection", 
               0, 100000, add.recommendation = FALSE, ylim=c(0, 1))
abline(h=1, col="red")
```

The plot demonstrates that even with 100,000 trees, perfect selection stability is not achievable. However, because the relationship between stability and tree count is known for this data set, the number of trees required to achieve specific levels of selection stability can be estimated using the `estimate_numtrees` function. For example:

```{r message=FALSE}
estimate_numtrees(optRF_result, measure="selection", for_stability = 0.9)
estimate_numtrees(optRF_result, measure="selection", for_stability = 0.95)
estimate_numtrees(optRF_result, measure="selection", for_stability = 0.99)
estimate_numtrees(optRF_result, measure="selection", for_stability = 0.999)
```

Increasing the target stability shows significant trade-offs in computation time. Achieving a stability of 0.9 requires approximately 19,658 trees and 35 seconds of computation time, while increasing the stability to 0.95 doubles the tree count and time. Achieving near-perfect stability of 0.999, however, would require over 10 million trees and almost 19 hours of computation, illustrating the exponential increase in computational demand as the stability target approaches 1.

It's essential to note that `opt_importance` estimates computation time based on the machine used to execute the function. The time required may vary on other machines. Additionally, since `ranger` supports parallel processing, computation time can be reduced by setting the `num.threads` argument, both in `ranger` and `opt_importance`.

Let's assume that after using the `opt_importance` function, we decide arbitrarily to use 500,000 trees for variable selection with random forest. Based on the results from the `estimate_numtrees` function, we know that this will lead to a selection stability slightly above 0.99. However, in order to publish the results together with the exact variable importance and selection stability, we can use the function `estimate_stability` with this specific number of trees: 

```{r message=FALSE}
estimate_stability(optRF_result, with_num.trees=500000)
```

The output confirms that with 500,000 trees, variable importance stability is approximately 0.9965, and selection stability is about 0.9903. These values provide a robust assessment of the reproducibility of results and reinforce the reliability of the analysis. Thus, the variable importance and selection stability should always be given when publishing variable selection based on random forest. 



## Customising the `opt_importance` function
This vignette demonstrated a specific application of the `opt_importance` function, assuming that only about 0.1% of the variables in the data set influence the response. Using this assumption, the number of trees was adjusted to enable stable variable selection. However, this step becomes unnecessary if a user already knows the exact number of variables to select. In such cases, the desired number of variables can simply be specified directly in the `alpha` argument of the `opt_importance` function.

In this example, the optimal number of trees was defined as the number of trees at which adding more trees would increase the variable importance stability by $10^{-6}$ or less. This threshold is arbitrary and can be customized using the `rec.thresh` argument to fit specific requirements.
Additionally, the `recommendation` argument offers further flexibility:

*   Set to `"importance"` or `"selection"` to optimize based on variable importance stability or selection stability, respectively.
*   Set to `"none"` to analyze the relationship between the number of trees and stability without receiving a specific recommendation. The user can then manually determine the optimal number of trees using the `estimate_numtrees` function.

Users can also adjust computational settings to tailor the `opt_importance` function to their needs:

*   Number of repetitions:
    Use the `number.repetitions` argument to control how many repetitions are performed. Fewer repetitions will reduce computation time but may lead to more variable results.

*   Specific tree values:
    The `num.trees_values` argument allows specifying particular numbers of trees to analyze. Smaller ranges or fewer values will speed up computation but might reduce the precision of the stability analysis.

    Recommendation: Avoid reducing these values unnecessarily, as doing so can compromise the accuracy of the recommendations provided by `opt_importance`.

Next to the arguments described here, the user can also add arguments from the `ranger` function to customise the application of `opt_prediction`. This gives the user the possibility to adjust the `num.threads` argument to allow for parallel computing or add specific values for hyperparameters in random forest such as `mtry`, `min.node.size`, or `sample.fraction`. 



## Conclusions
While random forest offers numerous advantages and is a powerful tool for estimating variable importance in a data set, it is essential to recognize its inherent non-deterministic nature. This characteristic can result in variability in variable importance estimates, even when analyzing the same data set multiple times. The stability of random forest outputs is heavily influenced by the number of trees used in the model. While increasing the number of trees enhances stability, it also leads to greater computational demands.

The optRF package provides an effective solution for addressing this trade-off by enabling the rapid modeling of the relationship between the number of trees and the stability of random forest. Through its opt_importance function, users can analyze this relationship for a specific data set and determine the optimal number of trees required to achieve desired stability levels while balancing computational efficiency.

Researchers are strongly encouraged to assess the stability of their chosen tree count to ensure reliable and consistent results when using random forest.  
